Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-05 Dail_Mad_ext.tex
plots from Jeff, and state-space description

One of the simplest state-space models can be described as follows.
% In their simplest form, state-space models include a Markovian model
%for the ecological process of interestTypically, the
%ecological
%process is described by a Markovian model that includes a
%deterministic description of population change as well as two random
%sources of variation for ``process variation''. The model for the
%Robservation process is typically phenomenological in the sense that
%observation error is not defined explicitly. As an example,
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and $X_t$ be
the observed data, which differs from $N_t$ due to observation error
($\zeta$). Temporal variation in $N_t$ is modeled using a
deterministic population growth model (e.g. exponential, logistic, etc
\dots) and stochastic model allowing for random process variation. A
simple state
space model that includes exponential growth and process variation
($\grave{o}$) could be written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
Here   is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

>>>>

One of the simplest state-space models can be described as follows.
% In their simplest form, state-space models include a Markovian model
%for the ecological process of interestTypically, the
%ecological
%process is described by a Markovian model that includes a
%deterministic description of population change as well as two random
%sources of variation for ``process variation''. The model for the
%Robservation process is typically phenomenological in the sense that
%observation error is not defined explicitly. As an example,
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and stochastic model allowing
for random process variation, which we denote $\grave{o}$ . As an
example, under exponential population growth, the model could be
written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
where $r$ is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.
